Supermartingales and Termination Analysis of Statistical PCF

### 2021-09-27 LICS21/SPCF.tex
Fix citations and references

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \cite{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \cite{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

>>>>

The language PPCF is a call-by-value (CBV) version of PCF with sampling of real numbers from the closed interval $[0,1]$ \citep{Ehrhard2018c,DBLP:journals/pacmpl/EhrhardPT18,DBLP:conf/esop/MakOPW21}.
%\changed[lo]{following \cite{DBLP:conf/esop/MakOPW21} except the score function (\Cref{rem:score})}. 
Types and terms are defined as follows, where $r$ is a real number, $x$ is a variable, $f : \mathbb{R}^n \to \mathbb{R}$ is any measurable function, and $\Gamma$ is an environment:
\begin{align*}
  \text{\em types } A, B &::= \textsf{R}  \mid  A \to B \\
  \text{\em values } V &::= \lambda x.M  \mid  \underline{r} \\
  \text{\em terms } M, N &::= V  \mid  x  \mid  M_1 \, M_2  \mid  \underline{f}(M_1,\dots ,M_n)  \mid  \tY \, M \\
   & \qquad \mid  \tif{M < 0}{N_1}{N_2}  \mid  \tsample
\end{align*}
The typing rules are standard (see \Cref{fig:typing rules}).
Terms are identified up to $\alpha$-equivalence, as usual. 
The set of all terms is denoted $\Lambda$, and the set of closed terms is denoted $\Lambda^0$.

@@

The reduction function is measurable, and the set of values is measurable, therefore the set of $s$ such that $M$ terminates at $s$ within $n$ steps is measurable for any $n$, therefore $\{s \mid \hbox{$M$ terminates at $s$}\}$ is measurable \citep{DBLP:conf/icfp/BorgstromLGS16,DBLP:conf/esop/MakOPW21}. 
We say that a term $M$ is \emph{almost surely terminating} (\emph{AST}) just if $\mu_{\entrosp}(\{s \mid \hbox{$M$ terminates at $s$}\}) = 1$.

### 2021-09-27 LICS21/antitone-proofs.tex
Moving some proofs back to main text: ongoing
Fix citations and references

\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \cite[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
\[
\underbrace{\big
(\tY \, \lambda f x .\ 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, \underline{10}
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \ln(n-1)$ works. The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

Many of the recent advances in the development of AST verification methods \cite{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \cite{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

There is an important formalism for constructing (non-random) higher-order recursive functions, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
%(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \cite{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
(order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
For example, the PPCF term in \Cref{ex:higher-order recursion} is not definable in PHORS.
%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
%cannot be expressed as PHORS.
%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}

A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \cite{DBLP:conf/lics/KobayashiLG19}.

@@

\antitoneSeemsComplete*

@@

The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.

>>>>

\subsection*{\Cref{ex:Fair-in-the-limit random walk} (Fair-in-the-limit random walk) \citep[\S 5.3]{DBLP:journals/pacmpl/McIverMKK18}}
\[
\underbrace{\big
(\tY \, \lambda f x .\ 
\tif{x \leq 0}{0}{f(x - 1) \oplus_{\frac{x}{2x+1}} f(x + 1)} \big)}_{\Xi} 
\, \underline{10}
\]
To construct an antitone ranking function, we solve the recurrence relation:
% \[
% \left\{
% \begin{array}{rll}
% z_0 &=& 0\\
% n \geq 1, \quad z_n &>& \dfrac{n}{2n + 1} \, z_{n-1} + \dfrac{n+1}{2n + 1} \, z_{n+1}.
% \end{array}
% \right.
% \]
\begin{align*}
z_0 &= 0\\
n \geq 1, \quad z_n &> \textstyle \frac{n}{2n + 1} \, z_{n-1} + \frac{n+1}{2n + 1} \, z_{n+1}.
\end{align*}
For $n > 2$, $z_n = \ln(n-1)$ works. The expected decrease is $\frac 1 2(\ln(1+\frac 1 {(n-1)^2-1}) - \frac 1 {2n+1} \ln(1 + \frac 2 {n-2}))$, and using the fact that $\frac x {x+1} \leq \ln (1+x) \leq x$ for $x > 0$, this is at least $\frac 1 2(\frac 1 {(n-1)^2} - \frac 1 {2n + 1} \frac 2 {n-2}) = \frac {n^2}{2(n-1)^2(2n+1)(n-2)}$, which (again for $n > 2$) is positive and antitone. For $n = 0, 1, 2$, we take $\epsilon$ to be 9/40 (the same as its value at 3), then set $z_2 = 2 \ln 2 - \ln 3 - 1, z_1 = 3 \ln 2 - 2 \ln 3 - 3, z_0 = 4 \ln 2 - 3 \ln 3 - 6$. Some of those values are negative, but it's still bounded below, so by adding a constant offset it can be corrected, and the term is AST.

@@

Many of the recent advances in the development of AST verification methods \citep{DBLP:conf/pldi/ChenH20,DBLP:conf/cav/ChakarovS13,DBLP:conf/popl/FioritiH15,DBLP:journals/pacmpl/McIverMKK18,DBLP:conf/aplas/HuangFC18,DBLP:conf/popl/ChatterjeeNZ17,DBLP:journals/pacmpl/AgrawalC018,DBLP:conf/cav/ChatterjeeFG16,DBLP:conf/lics/OlmedoKKM16,DBLP:journals/pacmpl/Huang0CG19} are concerned with loop-based programs.
We can view such loops as tail-recursive programs that are, in particular, \emph{affine recursive}, i.e., 
in each evaluation (or run) of the body of the recursion, recursive calls are made from at most one call site \cite[\S 4.1]{DBLP:journals/toplas/LagoG19}.
(Note that whether a program is affine recursive cannot be checked by just counting textual occurrences of variables.)
Termination analysis of \emph{non-affine recursive} probabilistic programs does not seem to have received much attention.
Methods such as those presented in \cite{DBLP:journals/toplas/LagoG19} are explicitly restricted to affine programs, and are unsound otherwise.
By contrast, many probabilistic programming languages allow for richer recursive structures \citep{DBLP:conf/pkdd/TolpinMW15,DBLP:conf/uai/GoodmanMRBT08,DBLP:journals/corr/MansinghkaSP14}.

@@

There is an important formalism for constructing (non-random) higher-order recursive functions, viz., \emph{higher-order recursion schemes} (HORS) (see e.g.~\cite{DBLP:conf/lics/Ong06,DBLP:conf/lics/Ong15}).
%(Incidentally the (sparse) ranking function method is just as applicable to the termination analysis of \emph{deterministic} PPCF programs.)
Recently \cite{DBLP:conf/lics/KobayashiLG19} have extended HORS to \emph{probabilistic higher-order recursion schemes} (PHORS), which are HORS augmented with probabilistic (binary) branching $\oplus_p$.
As HORS are in essence the $\lambda \tY$-calculus \citep{DBLP:conf/lics/Statman02} (i.e.~pure simply-typed lambda calculus with recursion, generated from a finite base type), PHORS are definable in PPCF:
(order-$n$) PHORS is encodable as (order-$n$) call-by-name PPCF, but the former is strictly less expressive (because the underlying HORS is not Turing complete). 
For example, the PPCF term in \Cref{ex:higher-order recursion} is not definable in PHORS.
%Some interesting PPCF terms such as \Cref{ex:non-affine with continuous distribution} %\refExample{complexExample} 
%cannot be expressed as PHORS.
%\akr{As I removed the referenced example, and I don't even know what a PHORS is, I've just removed this comment too.}

A relevant result here is that the AST problem is decidable for order-1 PHORS (by reduction to the PSPACE-hard solvability of finite systems of polynomial equations with real coefficients; see \cite{DBLP:journals/jacm/EtessamiY09}), but undecidable for PHORS with order greater than 1 \citep{DBLP:conf/lics/KobayashiLG19}.

@@

\iffalse
\antitoneSeemsComplete*

@@

The assumption that $(\mathcal F_n)_{n \geq 0}$ is the coarsest filtration to which $T$ is adapted is used in the equality between the third and fourth lines here. This condition is the main reason that this proof does not extend directly to a completeness result for antitone ranking functions.
\fi

### 2021-09-27 LICS21/antitone.tex
Fix citations and references

\iffalse
Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \cite{DBLP:conf/uai/GoodmanMRBT08}, Stan \cite{carpenter2017stan}, Anglican \cite{DBLP:conf/pkdd/TolpinMW15}, Gen \cite{cusumano-towner2019Gen}, Pyro \cite{bingham2019Pyro}, Edward \cite{tran2016edward} and Turing \cite{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.
\fi

>>>>

\iffalse
Sampling from continuous distributions is an essential feature of statistical probabilistic programming languages. (See e.g.~Church \citep{DBLP:conf/uai/GoodmanMRBT08}, Stan \citep{carpenter2017stan}, Anglican \citep{DBLP:conf/pkdd/TolpinMW15}, Gen \citep{cusumano-towner2019Gen}, Pyro \citep{bingham2019Pyro}, Edward \citep{tran2016edward} and Turing \citep{ge2018Turing}.)
Methods of proving AST of probabilistic computation have been developed for probabilistic programs with discrete distributions (see e.g.~\cite{DBLP:journals/toplas/LagoG19,DBLP:journals/jacm/KaminskiKMO18,DBLP:conf/lics/OlmedoKKM16,DBLP:conf/lics/KobayashiLG19,DBLP:conf/mfcs/KaminskiK15,DBLP:series/mcs/McIverM05}).
To our knowledge, the problem of proving AST of probabilistic functional programs with continuous distribution is new.
\fi

### 2021-09-27 LICS21/applications.tex
Fix citations and references

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \cite{kiselyov2016problems,Hur2015a}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
%\lo{It would be very interesting (and rewarding, I believe) to work out the details. There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}. SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
Selecting an appropriate addressing scheme is also important for variational inference \cite{paige2016automatic}.
%\cite[\S 6.2.2]{paige2016automatic}.

>>>>

It is possible to reduce this problem by labelling the samples in the trace by something other than the order in which they are used. If the labels of samples in the trace correspond closely to the roles that the samples play in the program execution, this will increase the efficiency of the inference algorithm. 
In \cite{wingate2011lightweight}, for example, something rather like a stack trace is used, although this addressing scheme turns out to have various problems \citep{kiselyov2016problems,Hur2015a}. 
Potential positions could also be used for a similar purpose, although in practice, a more compact representation than an entire skeletal reduction sequence would be necessary. 
%\lo{It would be very interesting (and rewarding, I believe) to work out the details. There is a formalisation of LMH for Idealised Anglican in \Cref{fnote:lmh}. SPCF is a much nicer probabilistic programming language (in purified / idealised form).}
Selecting an appropriate addressing scheme is also important for variational inference \citep{paige2016automatic}.
%\cite[\S 6.2.2]{paige2016automatic}.

### 2021-09-27 LICS21/confluent-proofs.tex
Move figures back to main text, and make alignment tweaks.
Removed \onecolumn and \twocolumn - seems to compile now
Fix citations and references

@@

\include{sim-diagram}
%The diagram is not actually in the figure environment because that doesn't work properly with it extending over several pages. The empty figure is just needed to get the label and caption.
\begin{figure}[htb]
\caption{Illustration of the $\sim_c$ rules. \label{sim-diagram}
}
\end{figure}

@@

  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$. Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first. Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \cite{Bar84,Takahashi1995}. This induction is in reverse, with $k$ decreasing from $n$ to $0$. If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$. If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired. In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.

@@

\begin{lemma} \label{lem:independentSamples}
For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
\end{lemma}
\begin{proof}
Suppose, to begin with, that $n = 0$. Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.

@@

\begin{align*}
& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
    & g_O(r,s'(O,f(O[r]))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
    & s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S\})\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O[r])) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O[r]))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
\end{align*}
\end{proof}
\clearpage

>>>>

\include{sim-diagram}

@@

@@

  Let this subsequence be $A \sim_\uparrow^n B \sim_\downarrow C$. Suppose for induction that $A \sim_\uparrow^k (D,\delta) \sim_\downarrow^* (E,\epsilon) \sim_\uparrow^* C$ for some $D, \delta, E, \epsilon$, and $0 \leq k \leq n$, where in the reduction sequence $D \to^* E$, for any pair of reductions at positions which aren't disjoint, the reduction at the innermost (greater) position occurs first. Reduction sequences with this property will be called ``parallel'', as it is analagous to the parallel reduction introduced by Tait and Martin-L\"of in their proof of the Church-Rosser theorem \citep{Bar84,Takahashi1995}. This induction is in reverse, with $k$ decreasing from $n$ to $0$. If $k = n$, simply set $(D,\delta) = B$ and $(E,\epsilon) = C$. If $k = 0$, then $A$ is related to $B$ by a sequence of $\sim_\downarrow$s then $\sim_\uparrow$s, as desired. In the intermediate steps, $k$ must be decreased by one, so a subsequence of one $\sim_\uparrow$ followed by a parallel sequence of $\sim_\downarrow$s must be replaced by a parallel sequence of $\sim_\downarrow$s followed by some $\sim_\uparrow$s.

@@

\begin{lemma} \label{lem:independentSamples}
For any initial term $M$, reduction strategy $f$ on $M$, natural number $n$, skeleton $N$ with $k$ holes, measurable set $T \subset \mathbb R ^ k$ and measurable set $S$ of samples in $I^{L_s(N)}$, $\mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) = \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)$
\end{lemma}

\begin{proof}
Suppose, to begin with, that $n = 0$. Either $\exists r \in T : N[r] = M$, in which case both sides of the equation are $\mu(S)$, or there is no such $r$, in which case both sides of the equation are 0.

@@

\begin{align*}
& \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in S : (M,s) \Rightarrow_f^n (N[r], s')\}) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\})\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in S : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}:\\
    & g_O(r,s'(O,f(O[r]))) \in T, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
     + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
     & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)}:\\
    & s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}:\\
    & g_O(r) \in T, s' \circ i(O[r] \to N[g_O(r)]) \in S, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    & \mu(\{s' \in I^{L_s(O)} \mid s'(O,f(O[r])) \in T_{h(O)}, s' \circ i(O[r] \to N[g_O(r,s'(O,f(O[r])))]) \in S\})\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\})\\
    & \mu({s' \in I^{L_s(O)} \mid s' \circ i(O[r] \to N[g_O(r)]) \in S}) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(T_{h(O)}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \prod_{j \neq h(O)} T_j, s' \in I^{L_s(O)} : s'(O,f(O[r])) \in T_{h(O)}, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \sum_{O \in R_s} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \mathbb R^{k-1}, s' \in I^{L_s(O)} : g_O(r,s'(O,f(O[r]))) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S)\\
    + & \sum_{O \in R_d} \mu(\{s \in I^{L_s(M)} \mid \exists r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}: g_O(r) \in T, (M,s) \Rightarrow_f^{n-1} (O[r], s')\}) \mu(S) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_s, r \in \mathbb R^{k-1}, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S)\\
    + & \mu(\{s \in I^{L_s(M)} \mid \exists O \in R_d, r \in \bigcup_l \mathbb R^l, s' \in I^{L_s(O)}, r' \in T, s'' \in I^{L_s(N)} : (M,s) \Rightarrow_f^{n-1} (O[r], s') \Rightarrow_f (N[r'], s'')\}) \mu(S) \\
= & \mu(\{s \in I^{L_s(M)} \mid \exists r \in T, s' \in I^{L_s(N)} : (M,s) \Rightarrow_f^n (N[r], s')\}) \mu(S)
\end{align*}
\end{proof}
